Recursive Solutions to First-Order Model Counting

### 2023-01-23 doc/paper/paper.tex
(Messy) preparation for submission

%File: anonymous-submission-latex-2023.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai23}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}

% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
% \usepackage{algorithm}
% \usepackage{algorithmic}

% These are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
% \usepackage{newfloat}
% \usepackage{listings}
% \DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
% \lstset{%
% 	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
% 	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
% 	aboveskip=0pt,belowskip=0pt,%
% 	showstringspaces=false,tabsize=2,breaklines=true}
% \floatstyle{ruled}
% \newfloat{listing}{tb}{lst}{}
% \floatname{listing}{Listing}

% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2023.1)
}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% Title

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\title{Recursive Solutions to First-Order Model Counting}
\author{Anonymous Submission}
\affiliations{}

\usetikzlibrary{cd}
\usetikzlibrary{shapes}

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

% TODO: when describing experimental results, say that a solution for the last
% (unsolved) instance can be found, but is too complicated to evaluate by hand.
% Add that the linear order axiom is also liftable. Cite the same paper in the
% introduction as well.

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting \citep{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory \citep{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics \citep{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the domain(s)
\citep{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be liftable
\citep{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, \citet{DBLP:conf/nips/Broeck11} showed that the class of all sentences of
FOL with up to two variables (denoted \FOtwo{}) is liftable. Then
\citet{DBLP:conf/pods/BeameBGS15} proved that there exists a sentence with three
variables for which FOMC is $\#\P_{1}$-complete (i.e., \FOthree{} is not
liftable). Since these two results came out, most of the research on (W)FOMC
focused on developing faster solutions for the \FOtwo{} fragment
\citep{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining new
liftable fragments. These fragments include \SFO{} and \SRU{}
\citep{DBLP:conf/nips/KazemiKBP16}, \Uone{} \citep{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \citep{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\citep{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the
empirical front, there are several open-source implementations of exact WFOMC
algorithms: \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving \citep{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C} \citep{DBLP:conf/kr/KazemiP16}. Approximate counting is supported
by \textsc{ApproxWFOMC} \citep{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \citep{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \citep{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is previous work
  \citep{DBLP:journals/jair/Kuzelka21,DBLP:conf/ijcai/Broeck16} showing how FOMC
  can compute factorials in principle, but an algorithm that can lift such an
  instance in practice does not exist.} We claim that this shortcoming is due to
the inability of these algorithms to construct recursive solutions. The topic of
recursion in the context of WFOMC has been studied before but in limited ways.
\Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data that
is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

@@

% search
% interpretation

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. We use lowercase Latin letters for predicates and constants, uppercase
Latin letters for variables, and uppercase Greek letters for domains. An
\emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all (finite)
domains. Initially, $\mathcal{D}$ contains all domains mentioned by the input
(W)FOMC instance. During compilation, new domains are added to $\mathcal{D}$ by
some of the compilation rules. Each such new domain is interpreted as a subset
of some element of $\mathcal{D}$.

@@

\begin{definition}[Clause]\label{def:clause}
  A \emph{clause}\footnote{\citet{DBLP:conf/ijcai/BroeckTMDR11} refer to clauses
    as c-clauses.} is a triple $c = (L, C, \delta_c)$, where $L$ is a set of
  literals, $C$ is a set of constraints, and $\delta_c$ is the domain map of
  $c$. Let $\Vars$ be the function that maps clauses and sets of either literals
  or constraints to the set of variables contained within. In particular,
  $\Vars(c) \coloneqq \Vars(L) \cup \Vars(C)$. \emph{Domain map}
  $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function that maps all
  variables in $c$ to their domains such that (s.t.) if $(X, Y) \in C$ for some
  variables $X$ and $Y$, then $\delta_c(X) = \delta_c(Y)$. For convenience, we
  sometimes write $\delta_c$ for the domain map of $c$ without unpacking $c$
  into its three constituents.
\end{definition}

@@

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by \citet{DBLP:conf/ijcai/BroeckTMDR11})
  is a set of clauses s.t.\ all constraints and atoms `type check' with respect
  to domains.
\end{definition}

@@

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by \citet{DBLP:conf/ijcai/BroeckTMDR11}. First, we decouple
variable-to-domain assignments from constraints and move them to a separate
function $\delta_{c}$ in \cref{def:clause}. Formalising these assignments as a
function unveils the (previously implicit) assumption that each variable must be
assigned to a domain. Second, while \citet{DBLP:conf/ijcai/BroeckTMDR11} allow
for equality constraints and constraints of the form $X \not\in \Delta$ for some
variable $X$ and domain $\Delta$, we exclude such constraints simply because
they are not needed.

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|\Gamma| = |\Delta| = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $p$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $\phi$ (when $|\Gamma| = |\Delta| = 2$)
  is 7. Incidentally, the FOMC of $\phi$ counts partial injections. We will
  continue to use the problem of counting partial injections (and the formula
  from \cref{example:first} specifically) as the main running example throughout
  the paper.
\end{example}
% TODO: It might be useful to give one or two readable examples with interesting
% model counts. For example Px v Qx over domain n has 3^n/4^n solutions. I think
% The Frends smokers example has a more complicated solution you can find in one
% of the papers. It might motivate the framework generally speaking better
% although of course you can continue to use the same running example for the
% rest of your paper. Such examples are especially useful for readers who are
% not familiar but still need to get excited by the problem

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit some parameters as well as nodes whose only arithmetic
    effect is multiplication by one.}\label{fig:examplefcg}
\end{figure}
% Some of these nodes play an important role in the weighted version of the
% problem, whereas others are remnants of the interaction between compilation
% rules and the way in which \textsc{ForcLift} handles existential quantifiers.

@@

% \footnote{Imposing an ordering on outgoing edges is just a limited version of
% edge labelling.}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges. Node
labels consist of two parts: the \emph{type} and the \emph{parameters}. The type
of a node determines its out-degree.

@@

Along with many node types defined previously
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}, \textsc{Crane}
uses three more:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
Similarly to \citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG
that has a node with the label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors. See \cref{fig:examplefcg} for an
example FCG\@. Its source node has out-degree 1, label
$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule does not apply to the formula.

>>>>

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in

\usepackage{kr}

% Use the postscript times font!
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
% \usepackage{algorithm}
% \usepackage{algorithmic}
\urlstyle{same}

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usetikzlibrary{cd}
\usetikzlibrary{shapes}

% the following package is optional:
%\usepackage{latexsym}

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\pdfinfo{
/TemplateVersion (KR.2022.0, KR.2023.0)
}

\title{Recursive Solutions to First-Order Model Counting}

% TODO: double check how the university names should be formatted and the emails
% TODO: read the formatting instructions and their LaTeX source
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore\\
$^2$University of Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting \cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals \cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \cite{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \cite{DBLP:journals/ml/KimmigMG15} and statistical
relational \cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory \cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics \cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

% TODO: have ~ in front of every \cite
% TODO: fix doom emacs warnings
Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the domain(s)
\cite{jaeger2012liftability}. Over more than a decade, many classes of instances
were shown to be liftable
\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment \cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and \SRU{}
\cite{DBLP:conf/nips/KazemiKBP16}, \Uone{} \cite{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the empirical
front, there are several open-source implementations of exact WFOMC algorithms:
\textsc{ForcLift} \cite{DBLP:conf/ijcai/BroeckTMDR11}, probabilistic theorem
proving \cite{DBLP:journals/cacm/GogateD16}, and \textsc{L2C}
\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{ApproxWFOMC} \cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \cite{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is previous work
  \cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/ijcai/Broeck16} showing how FOMC
  can compute factorials in principle, but an algorithm that can lift such an
  instance in practice does not exist.} We claim that this shortcoming is due to
the inability of these algorithms to construct recursive solutions. The topic of
recursion in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can
translate sentences in a variant of function-free many-sorted FOL with equality
to this internal format. We use lowercase Latin letters for predicates and
constants, uppercase Latin letters for variables, and uppercase Greek letters
for domains. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and
terms $t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all (finite)
domains. Initially, $\mathcal{D}$ contains all domains mentioned by the input
(W)FOMC instance. During compilation, new domains are added to $\mathcal{D}$ by
some of the compilation rules. Each such new domain is interpreted as a subset
of some element of $\mathcal{D}$.

@@

\begin{definition}[Clause]\label{def:clause}
  A \emph{clause}\footnote{Van den Broeck et
    al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} refer to clauses as c-clauses.}
  is a triple $c = (L, C, \delta_c)$, where $L$ is a set of literals, $C$ is a
  set of constraints, and $\delta_c$ is the domain map of $c$. Let $\Vars$ be
  the function that maps clauses and sets of either literals or constraints to
  the set of variables contained within. In particular,
  $\Vars(c) \coloneqq \Vars(L) \cup \Vars(C)$. \emph{Domain map}
  $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function that maps all
  variables in $c$ to their domains such that (s.t.) if $(X, Y) \in C$ for some
  variables $X$ and $Y$, then $\delta_c(X) = \delta_c(Y)$. For convenience, we
  sometimes write $\delta_c$ for the domain map of $c$ without unpacking $c$
  into its three constituents.
\end{definition}

@@

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by Van den Broeck et
  al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}) is a set of clauses s.t.\ all
  constraints and atoms `type check' with respect to domains.
\end{definition}

@@

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Formalising these
assignments as a function unveils the (previously implicit) assumption that each
variable must be assigned to a domain. Second, while Van den Broeck et
al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} allow for equality constraints and
constraints of the form $X \not\in \Delta$ for some variable $X$ and domain
$\Delta$, we exclude such constraints simply because they are not needed.

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|\Gamma| = |\Delta| = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $p$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $\phi$ (when $|\Gamma| = |\Delta| = 2$)
  is 7. Incidentally, the FOMC of $\phi$ counts partial injections. We will
  continue to use the problem of counting partial injections (and the formula
  from \cref{example:first} specifically) as the main running example throughout
  the paper.
\end{example}
% TODO: It might be useful to give one or two readable examples with interesting
% model counts. For example Px v Qx over domain n has 3^n solutions. I think
% The Frends smokers example has a more complicated solution you can find in one
% of the papers. It might motivate the framework generally speaking better
% although of course you can continue to use the same running example for the
% rest of your paper. Such examples are especially useful for readers who are
% not familiar but still need to get excited by the problem

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    Van den Broeck et al.~\protect\shortcite{DBLP:conf/ijcai/BroeckTMDR11} for
    the descriptions of these node types. Here we omit some parameters as well
    as nodes whose only arithmetic effect is multiplication by
    one.}\label{fig:examplefcg}
\end{figure}

@@

A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges. Node
labels consist of two parts: the \emph{type} and the \emph{parameters}. The type
of a node determines its out-degree.

@@

Along with many node types defined previously
\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}, \textsc{Crane}
uses three more:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
Similarly to Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}, we
write $T_p$ for an FCG that has a node with the label $T_p$ (i.e., type $T$ and
parameter(s) $p$) and $\star$'s as all of its direct successors. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\cite{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule does not apply to the formula.
